Astr 513: Final Project

### 2016-05-02 Final_Project/ms.tex
Edited SN section, added Betoule cov matrix description.

\section{Type \rom{1}a Supernovae}
\label{sec:SNIa}
\EW{You guys might want to read this over and tell me if you think I should
  change/add anything.}
A type \rom{1}a supernova (\sna{}) is a supernova with very little hydrogen or
helium absorption. It is believed that the progenitor for these special
supernovae are white dwarfs that are part of a binary system. These white
dwarfs accrete mass from their companion stars or merge with a companion white
dwarf or neutron star, causing them to exceed the Chandrasekhar limit
($1.44 M_\odot$). When this happens, carbon and oxygen fusion begins in the
core, and the temperature increases. However, the white dwarf does not expand
and cool as a main sequence star would, since it is supported by degeneracy
pressure rather than thermal, so the core experiences a runaway fusion
reaction, releasing $\sim 10^{44}$ J of energy
\citep[see, e.g.,][chap. 18.4]{ryden2010}, which is enough to overcome the
gravitational energy and the star explodes. Because all \snea{} are caused by a
similar reaction, they all have similar luminosities, and more importantly, the
light curves all have the same shape: a sharp maximum followed by a smooth,
gradual decay. Studying nearby \snea{} has shown that the rate of decline of
the light curve is correlated to the peak luminosity, and this correlation has
been calibrated using Cephied variable stars located in the same galaxies as
nearby \sna{}. With this information, we can estimate peak luminosity of an
\sna{} from its light curve, thereby providing us with a standard candle with
which to measure distance. They were first used in this way by the
Cal\'{a}n/Tololo Supernova Survey, and it was using \snea{} that
\citet{riess1998} and \citet{perlmutter1999} were able to discover the
accelerating expansion rate of the Universe for which three of the authors
later won the Nobel Prize in physics.

@@

\subsection{Covariances}
\label{sec:covariances}
\LC{Wrote about Rest's cov matrix. Still need description of Betoule's.}

@@

@@

\section{Distance Modulus Model}
\label{sec:model}
\LC{Need more here but I wrote most of the relevant eqns}
The main objective of \citet{betoule2014} and \citet{rest2014} is 
to place constraints on cosmological parameters $\Omega_M$, $\Omega_\Lambda$, 
and $w$. In order to so, we must examine the relationship between distance 
modulus and cosmology. The distance modulus $\mu$ is determined from 
the luminosity distance $d_L$ by 
\begin{equation}
\label{eq:mu_dL}
\mu=5 \text{log}(d_L)+25. 
\end{equation}
The luminosity distance depends on the cosmology by 
\begin{equation}
\label{eq:luminosityDistance}
d_L(z; w, \Omega_M, \Omega_\Lambda, H_0)= (1+z)|\Omega_k|^{-1/2}\mathcal{S}_{k}\displaystyle\left[c|\Omega_k|^{1/2}\int_{0}^{z}\frac{dz'}{H(z')}\right],
\end{equation}
where $\Omega_k \equiv 1-\Omega_M-\Omega_{\Lambda}$. The function 
$\mathcal{S}_{k}$ depends on the curvature density $\Omega_k$ in that 
\[
    \mathcal{S}_{k}= 
\begin{cases}
    \text{sin}(x),&  \Omega_k < 0\\
    x,& \Omega_k = 0\\
    \text{sinh}(x),& \Omega_k > 0
\end{cases}
\]
and the function $H(z)$ is 
\begin{equation}
\label{eq:hubbleFunction}
H(z)= H_0\displaystyle\left[\Omega_M(1+z)^3+\Omega_{\Lambda}(1+z)^{3(1+w)}+\Omega_k(1+z)^2\right]^{1/2}.
\end{equation}
Three (?) cosmologies are of interest to us: 
\LC{Not sure which ones we're interested in. $\Lambda CDM$, $\Omega_M=1$, ?}
\TM{I think there are only two, one with $\Omega_k=0$ and one with it
  being unconstrained.}

>>>>

\section{Type \rom{1}a Supernovae}
\label{sec:SNIa}
\EW{Should I cite the Nobel Prize information document where I found out about 
the stuff in the last few sentences? I'm not entirely sure what that would be considered...}
A type \rom{1}a supernova (\sna{}) is a supernova with very little hydrogen or
helium absorption. It is believed that the progenitor for these special
supernovae are white dwarfs that are part of a binary system. These white
dwarfs accrete mass from their companion stars or merge with a companion white
dwarf or neutron star, causing them to exceed the Chandrasekhar limit
($1.44 M_\odot$). When this happens, carbon and oxygen fusion begins in the
core, and the temperature increases. However, the white dwarf does not expand
and cool as a main sequence star would, since it is supported by degeneracy
pressure rather than thermal, so the core experiences a runaway fusion
reaction, releasing $\sim 10^{44}$ J of energy \citep[see, e.g.,][chap. 18.4]{ryden2010}, which is enough to overcome the gravitational energy and the star explodes. Because all \snea{} are caused by a similar reaction, they all have similar luminosities, and more importantly, the light curves all have the same shape: a sharp maximum followed by a smooth, gradual decay. Studying nearby \snea{} has shown that the rate of decline of the light curve is correlated to the peak luminosity, and this correlation has been calibrated using Cephied variable stars located in the same galaxies as nearby \sna{}. With this information, we can estimate peak luminosity of an \sna{} from its light curve, thereby providing us with a standard candle with which to measure distance. They were first used in this way in the 1990s by two independent teams, the Supernova Cosmology Project and the Cal\'{a}n/Tololo Supernova Survey. It was using distances from \snea{} from these two projects that \citet{riess1998} and \citet{perlmutter1999} were able to discover the accelerating expansion rate of the Universe in 1998. It was for this work that Saul Perlmutter, Brian P. Schmidt, and Adam G. Reiss won the Nobel Prize in physics in 2011.

@@

\subsection{Covariances}
\label{sec:covariances}
\LC{Wrote about Rest's cov matrix. Still need description of Betoule's.}
\EW{Wrote about Beoule's covariance matrix.}

@@

\citet{betoule2014} choose to first calculate the covariance matrix $\mathbf{C}_{\mathbf{\eta}}$ of the 3 light curve fit-parameters $\mathbf{\eta}$ by adding the individual covariance matrices they get from statistical errors propagated from light-curve fit uncertainties and 7 sources of systematic uncertainties they consider from their photometric calibartion, light-curve model uncertainty, bias correction, a mass step uncertainty introduced when correcting magnitudes due to host mass, Galactic extinction, peculiar velocities, and contamination by non-\sna{}. They then propagate these errors to the distance modulus and use three terms to account for uncertainty in cosmological redshift by the equation
\begin{equation}
\label{eq:BetouleCovMatrix}
\mathbf{C} = \mathbf{A} \mathbf{C}_\mathbf{\eta} \mathbf{A}^\dag + \mathrm{diag}\left(\frac{5 \sigma_z}{z \log 10}\right)^2 + \mathrm{diag}\left(\sigma_{\mathrm{lens}}^2\right) + \mathrm{diag}\left(\sigma_{\mathrm{coh}}^2\right),
\end{equation}
where $\mathbf{A}$ is the matrix of coefficients in the distance modulus, and redshift uncertainty is accounted for in the first diagonal term due to peculiar velocities, the second due to magnitude variation from gravitational lensing, and the third from intrinsic \sn{} magnitude variation not included in other terms. They use $c \sigma_z = 150 \mathrm{km}\, \mathrm{s}^{-1}$, $\sigma_{\mathrm{lens}} = 0.055 \times z$, and a value of $\sigma_{\mathrm{coh}}$ for each survey found using a restricted log-likelihood technique. The elements of the covariance matrix available online are calculated on the redshift control points $z_b$ as described in \S \ref{sec:betoule}.

@@

\section{Distance Modulus Model}
\label{sec:model}
\LC{Need more here but I wrote most of the relevant eqns}
\EW{Do we want/need citations for the equations here?}
The main objective of \citet{betoule2014} and \citet{rest2014} is 
to place constraints on cosmological parameters $\Omega_M$, $\Omega_\Lambda$, 
and $w$. In order to so, we must examine the relationship between distance 
modulus and cosmology. The distance modulus $\mu$ is determined from 
the luminosity distance $d_L$ by 
\begin{equation}
\label{eq:mu_dL}
\mu=5 \text{log}(d_L)+25. 
\end{equation}
The luminosity distance depends on the cosmology by 
\begin{equation}
\label{eq:luminosityDistance}
d_L(z; w, \Omega_M, \Omega_\Lambda, H_0)= (1+z)|\Omega_k|^{-1/2}\mathcal{S}_{k}\displaystyle\left[c|\Omega_k|^{1/2}\int_{0}^{z}\frac{dz'}{H(z')}\right],
\end{equation}
where $\Omega_k \equiv 1-\Omega_M-\Omega_{\Lambda}$. The function 
$\mathcal{S}_{k}$ depends on the curvature density $\Omega_k$ in that 
\[
    \mathcal{S}_{k}= 
\begin{cases}
    \text{sin}(x),&  \Omega_k < 0\\
    x,& \Omega_k = 0\\
    \text{sinh}(x),& \Omega_k > 0
\end{cases}
\]
and the function $H(z)$ is 
\begin{equation}
\label{eq:hubbleFunction}
H(z)= H_0\displaystyle\left[\Omega_M(1+z)^3+\Omega_{\Lambda}(1+z)^{3(1+w)}+\Omega_k(1+z)^2\right]^{1/2}.
\end{equation}
Three (?) cosmologies are of interest to us: 
\LC{Not sure which ones we're interested in. $\Lambda CDM$, $\Omega_M=1$, ?}
\TM{I think there are only two, one with $\Omega_k=0$ and one with it
  being unconstrained.}
